A sheaf--function dictionary for algebraic tori over local fields

### 2013-08-18 shanghai.tex
Moved the lemmas in the section on finite quasicharacter sheaves.

We put a tensor category structure on $C(G)$ 
and a rigid monoidal category structure on $C_1(G)$ as follows.
Let $(f,\psi)$ and $(f',\psi')$ be objects in $C(G)$. 
The category of isogenies to $G$ admits products: 
write $f\times_G f' : H\times_G H' \to G$ for the product of $f : H\to G$ and $f' : H'\to G$. 
Since $\ker(f\times_G f') \iso (\ker f)\times (\ker f')$, canonically,
the tensor product of the representations $\psi : \ker f\to \Aut(V)$ and $\psi' : \ker f'\to \Aut(V')$ 
is the representation $\psi \otimes\psi' :  \ker(f\times_G f') \to \Aut(V\otimes_{\EE} V')$. 
This allows us to set $(f,\psi)\otimes(f',\psi') \ceq (f\times_G f' ,\psi\otimes\psi')$. 
The same fact makes sense of $(f,\psi)\oplus(f',\psi')\ceq (f\times_G f' ,\psi\oplus\psi')$. 
The dual of $(f,\psi)$ is $(f,\psi)^\vee \ceq (f,\psi^\vee)$ 
where $\psi^\vee : \ker f \to \Aut(V^\vee)$ and $V^\vee = \Hom_\text{vec}(V,\EE)$. 
The verification that these operations make $C(G)$ a tensor category is straightforward.
Likewise, $C(G)[W^{-1}]$ is a tensor category.
Let $C_1(G)$ be the rigid monoidal full subcategory of $C(G)$ consisting of objects
$(f,\psi)$ such that $\psi$ is a one-dimensional representation.

@@

Although this section does not mention quasicharacter sheaves, 
we will need Proposition~\ref{prop:X} to prove our main result on quasicharacter sheaves, 
Theorem~\ref{thm:snake}.

@@

%\input{snake}
\begin{theorem}\label{thm:snake}
  If $G$ is a smooth abelian group scheme, locally of finite type over
  $\Fq$, and if the abelian group $\pi_0(G)(\bFq)$ (the geometric
  component group) is finitely generated, then the trace of Frobenius
  \[
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
  is an isomorphism of groups.
\end{theorem}

@@

We will also need the Greenberg functor, 
introduced in \cite{greenberg:61} and \cite{greenberg:63a}, 
as a functor from the category of schemes locally of finite type 
over an Artinian local ring with perfect residue field $k$ 
to the category of schemes locally of finite type over $k$. 
This functor respects open immersions, closed immersions, 
etale morphisms and smooth morphisms. 
Besides Greenberg's original papers, 
see \cite{bosch-lutkebohmert-reynaud:NeronModels}*{Ch. 9, \S 6} 
and the first few sections of \cite{stasinski:12a} 
for background on the Greenberg functor. 
Let
\[
%\begin{tikzcd}
\Gr_d : \left( \text{Schemes} \atop \text{lft over $\OK$}\right) \to \left( \text{Schemes} \atop \text{lft over $\Fq$}\right)
%\end{tikzcd}
\]
be the functor produced by precomposing pull-back along $\OK \to \OK/\pK^d$ with the Greenberg functor.
The surjective ring homomorphism $\OK/\pK^{d+1} \to \OK/\pK^d$ determines a natural transformation $P_{d} : \Gr_{d+1} \to \Gr_d$. 
%\cite{bosch-lutkebohmert-reynaud:NeronModels}*{p. 276}.

For any positive integer $d$ set $\GN{T}_d = \Gr_d(\TT)$ and $f_d = P_d(\TT)$.
The fundamental properties of $f_d : \GN{T}_{d+1} \to \GN{T}_d$ that we need are:
 \begin{enumerate}
   \labitem{(GN.0)}{dFq} $\GN{T}_d(\Fq) = \TT(\OK/\pK^d)$, canonically.
  \labitem{(GN.1)}{GNd} $\GN{T}_d$ is a smooth, commutative group scheme,
  locally of finite type over $\Fq$, and the component group of $\GN{T}_d$ is
  geometrically finitely generated. (In fact, $\pi_0(\GN{T}_d) = \pi_0(\TT) \times_{\Spec{\OK}} \Spec{\Fq}$,
  and the latter is finitely generated by \cite{xarles:93a}*{2.18}.)
  \labitem{(GN.2)}{truncation}
  The morphism $f_d : \GN{T}_{d+1} \to \GN{T}_d$ is a surjective morphism of group schemes.
  \end{enumerate}

@@

By Proposition~\ref{prop:pullback}, the surjective morphism $f_d : \GN{T}_{d+1}\to \GN{T}_d$ of group schemes induces a functor
\[
f_d^* : \QC(\GN{T}_d) \to \QC(\GN{T}_{d+1}).
\]
\begin{proposition}
The functor $f_d^* : \QC(\GN{T}_d) \to \QC(\GN{T}_{d+1})$ is full and faithful.
\end{proposition}

\begin{proof}
\Clifton{Under construction}
Base change along $\OK/\pK^{d+1} \to \OK/\pK^d$ determines a closed immersion $\TT_{d+1} \hookrightarrow \TT_d$, though not a morphism of group schemes. Applying the Greenberg functor produces a closed immersion $i_d : \GN{T}_{d} \hookrightarrow \GN{T}_{d+1}$. Although $i_d$ is not a morphism of group schemes, it is a section of $f_d : \GN{T}_{d+1}\to \GN{T}_d$.
\end{proof}

Let $\QC(\GN{T})$ be the category of ind-objects from the full subcategories $\QC(\GN{T}_d)$. 
Objects in $\QC(\GN{T})$ are equivalence classes of objects from $\QC(\GN{T}_d)$ under the equivalence relation that identifies 
$\qcs{L}_1\in \QC(\GN{T}_{d_1})$ with $\qcs{L}_2\in \QC(\GN{T}_{d_2})$ 
if $f_1^*\qcs{L}_1 = f_2^*\qcs{L}_2$  in $\QC(\GN{T}_{\max\{d_1,d_2\}})$,
where $f_1 : \GN{T}_{\max\{d_1,d_2\}} \to \GN{T}_{d_1}$ 
(resp. $f_2 : \GN{T}_{\max\{d_1,d_2\}} \to \GN{T}_{d_2}$) 
is obtained by composing the maps $f_{d_1}, \cdots , f_{\max\{d_1,d_2\}}$ 
(resp. $f_{d_2}, \cdots , f_{\max\{d_1,d_2\}}$).
A map $\alpha : \qcs{F} \to \qcs{F}'$

>>>>

We put a tensor category structure on $C(G)$ 
and a rigid monoidal category structure on $C_1(G)$ as follows.
Let $(f,\psi)$ and $(f',\psi')$ be objects in $C(G)$. 
The category of isogenies to $G$ admits products: 
write $f\times_G f' : H\times_G H' \to G$ for the product of $f : H\to G$ and $f' : H'\to G$. 
Since $\ker(f\times_G f') \iso (\ker f)\times (\ker f')$, canonically,
the tensor product of the representations $\psi : \ker f\to \Aut(V)$ and $\psi' : \ker f'\to \Aut(V')$ 
is the representation $\psi \otimes\psi' :  \ker(f\times_G f') \to \Aut(V\otimes_{\EE} V')$. 
This allows us to set $(f,\psi)\otimes(f',\psi') \ceq (f\times_G f' ,\psi\otimes\psi')$. 
The same fact makes sense of $(f,\psi)\oplus(f',\psi')\ceq (f\times_G f' ,\psi\oplus\psi')$. 
The dual of $(f,\psi)$ is $(f,\psi)^\vee \ceq (f,\psi^\vee)$ 
where $\psi^\vee : \ker f \to \Aut(V^\vee)$ and $V^\vee = \Hom_\text{vec}(V,\EE)$. 
The verification that these operations make $C(G)$ a tensor category is straightforward.
Likewise, $C(G)[W^{-1}]$ is a tensor category.

Let $C_1(G)$ be the rigid monoidal full subcategory of $C(G)$ consisting of objects
$(f,\psi)$ such that $\psi$ is a one-dimensional representation.

@@

\Clifton{Now this section does mention quasicharacter sheaves.}
Although this section does not mention quasicharacter sheaves, 
we will need Proposition~\ref{prop:X} to prove our main result on quasicharacter sheaves, 
Theorem~\ref{thm:snake}.

@@

%\input{snake}
\begin{theorem}\label{thm:snake}
  If $G$ is a smooth commutative group scheme locally of finite type over
  $\Fq$ and if the abelian group $\pi_0(G)(\bFq)$ (the geometric
  component group) is finitely generated then the trace of Frobenius
  \[
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
  is an isomorphism of groups.
\end{theorem}

@@

We will also need the Greenberg functor, 
introduced in \cite{greenberg:61} and \cite{greenberg:63a}, 
as a functor from the category of schemes locally of finite type 
over an Artinian local ring with perfect residue field $k$ 
to the category of schemes locally of finite type over $k$. 
This functor respects open immersions, closed immersions, 
etale morphisms and smooth morphisms. 
Besides Greenberg's original papers, 
see \cite{bosch-lutkebohmert-reynaud:NeronModels}*{Ch. 9, \S 6} 
and the first few sections of \cite{stasinski:12a} 
for background on the Greenberg functor.

For any positive integer $d$, let
\[
%\begin{tikzcd}
\Gr_d : \left( \text{Schemes} \atop \text{lft over $\OK$}\right) \to \left( \text{Schemes} \atop \text{lft over $\Fq$}\right)
%\end{tikzcd}
\]
be the functor produced by precomposing pull-back along $\OK \to \OK/\pK^d$ with the Greenberg functor.
The surjective ring homomorphism $\OK/\pK^{d+1} \to \OK/\pK^d$ determines a natural transformation $P_{d} : \Gr_{d+1} \to \Gr_d$. 
%\cite{bosch-lutkebohmert-reynaud:NeronModels}*{p. 276}.
Set $\GN{T}_d = \Gr_d(\TT)$ and $f_d = P_d(\TT)$.
%
The fundamental properties of $f_d : \GN{T}_{d+1} \to \GN{T}_d$ that we need are:
 \begin{enumerate}
   \labitem{(GN.0)}{dFq} $\GN{T}_d(\Fq) = \TT(\OK/\pK^d)$, canonically.
  \labitem{(GN.1)}{GNd} $\GN{T}_d$ is a smooth, commutative group scheme,
  locally of finite type over $\Fq$, and the component group of $\GN{T}_d$ is
  geometrically finitely generated. (In fact, $\pi_0(\GN{T}_d) = \pi_0(\TT) \times_{\Spec{\OK}} \Spec{\Fq}$,
  and the latter is finitely generated by \cite{xarles:93a}*{2.18}.)
  \labitem{(GN.2)}{truncation}
  The morphism $f_d : \GN{T}_{d+1} \to \GN{T}_d$ is a surjective morphism of group schemes.
  \end{enumerate}

@@

By Proposition~\ref{prop:pullback}, the surjective morphism $f_d : \GN{T}_{d+1}\to \GN{T}_d$ of group schemes induces a functor
\[
f_d^* : \QC(\GN{T}_d) \to \QC(\GN{T}_{d+1}).
\]

\begin{proposition}
The functor $f_d^* : \QC(\GN{T}_d) \to \QC(\GN{T}_{d+1})$ is full and faithful.
\end{proposition}

\begin{proof}
It suffices to find a morphism of schemes $i_d : \GN{T}_{d}\hookrightarrow \GN{T}_{d+1}$ such that $f_d\circ i_d = \id_{\GN{T}_d}$. Base change along $\OK/\pK^{d+1} \to \OK/\pK^d$ determines a closed immersion $\TT_{d+1} \hookrightarrow \TT_d$, though not a morphism of group schemes. Applying the Greenberg functor produces a closed immersion $i_d : \GN{T}_{d} \hookrightarrow \GN{T}_{d+1}$. Although $i_d$ is not a morphism of group schemes, it is a section of $f_d : \GN{T}_{d+1}\to \GN{T}_d$.
\end{proof}

Let $\QC(\GN{T})$ be the category of ind-objects built from the categories $\QC(\GN{T}_d)$; we refer to this as the \emph{category of quasicharacters for $T$}. This means a quasicharacter sheaf for $T$ is an equivalence class of quasicharacters on $\GN{T}_d$, for various $d$, under the relation that identifies $\qcs{L}\in \QC(\GN{T}_{d})$ with $f_d^*\qcs{L}\in \QC(\GN{T}_{d})$.
For each $\qcs{F}\in \QC(\GN{T})$ there is a unique least positive integer $d$ and a unique $\qcs{L}\in \GN{T}_d$ such that $\qcs{F} = [\qcs{L}]$; we refer to $d$ as the \emph{level} of $\qcs{F}$.
A map $\alpha : \qcs{F} \to \qcs{F}'$ in $\QC(\GN{T})$
